A tutorial on the model-dependent analysis of neutron and X-ray reflectometry

### 2020-07-23 refl_maths/paper.tex
fix fraction
Add section on ill posedness and the phase problem

@@

\author{Andrew R. McCluskey}
  \email{andrew.mccluskey@diamond.ac.uk}
  \email{a.r.mccluskey@bath.ac.uk}
  \affiliation{Diamond Light Source, Rutherford Appleton Laboratory, Harwell Science and Innovation Campus, Didcot, OX11 0DE, UK}
  \affiliation{Department of Chemistry, University of Bath, Claverton Down, Bath BA2 7AY, UK}
\author{Andrew R. J. Nelson}
  \affiliation{Bragg Institute, Australian Nuclear Science and Technology Organisation, Locked Bag 2001, Kirrawee DC, NSW 2232, Australia}
\author{Tim Snow}
  \affiliation{Diamond Light Source, Rutherford Appleton Laboratory, Harwell Science and Innovation Campus, Didcot, OX11 0DE, UK}
  \affiliation{School of Chemistry, University of Bristol, Bristol, BS8 1TS, UK}

@@

Here, we will briefly discuss the introduction of a simple Gaussian resolution function with a constant $\sfrac{\text{d}q}{q}$. 
The Gaussian function is centred on a given wavevector, $q$, and has a width of at half maximum of $\text{d}q$.
For each $q$-value in the unsmeared data, a Gaussian function is produced which is convolved with the unsmeared to produce data that includes some description of the instrumental resolution.
Typically, this convolvutional smearing is included in the reading of the experimental data. 
However, the resolution function used by a given software package may not be an accurate description of that from the instrument, therefore, we recommend discussing this with the members of instrument staff. 
More information about the resolution function can be found in the work of van Well and Fredrikze \cite{vanwell_resolution_2005} and Nelson and Dewhurst \cite{nelson_towards_2013,nelson_towards_2014}

>>>>

\newcommand{\rmd}{\rm{d}}
\newcommand{\diff}[2]{\frac{\rmd#1}{\rmd#2}}
\newcommand{\acf}{\rm{ACF}}
\newcommand{\angstrom}{\rm{\AA}}
\newcommand{\scinot}[2]{ {#1} \times 10^{#2} }

@@

\author{Andrew R. McCluskey}
  \email{andrew.mccluskey@diamond.ac.uk}
  \email{a.r.mccluskey@bath.ac.uk}
  \affiliation{Diamond Light Source, Rutherford Appleton Laboratory, Harwell Science and Innovation Campus, Didcot, OX11 0DE, UK}
  \affiliation{Department of Chemistry, University of Bath, Claverton Down, Bath BA2 7AY, UK}
\author{Andrew R. J. Nelson}
  \affiliation{Bragg Institute, Australian Nuclear Science and Technology Organisation, Locked Bag 2001, Kirrawee DC, NSW 2232, Australia}
\author{Tim Snow}
  \affiliation{Diamond Light Source, Rutherford Appleton Laboratory, Harwell Science and Innovation Campus, Didcot, OX11 0DE, UK}
  \affiliation{School of Chemistry, University of Bristol, Bristol, BS8 1TS, UK}
\author{Juan M. Carmona Loaiza}
  \affiliation{J\"ulich Centre for Neutron Science (JCNS) at Heinz Maier-Leibnitz Zentrum (MLZ).
	Lichtenbergstraße 1, 85748 Garching, Germany.}

@@

Here, we will briefly discuss the introduction of a simple Gaussian resolution function with a constant $\frac{\text{d}q}{q}$. 
The Gaussian function is centred on a given wavevector, $q$, and has a width of at half maximum of $\text{d}q$.
For each $q$-value in the unsmeared data, a Gaussian function is produced which is convolved with the unsmeared to produce data that includes some description of the instrumental resolution.
Typically, this convolvutional smearing is included in the reading of the experimental data. 
However, the resolution function used by a given software package may not be an accurate description of that from the instrument, therefore, we recommend discussing this with the members of instrument staff. 
More information about the resolution function can be found in the work of van Well and Fredrikze \cite{vanwell_resolution_2005} and Nelson and Dewhurst \cite{nelson_towards_2013,nelson_towards_2014}

%%%%
\section{Ill posedness - the phase problem}

Figure \ref{fig:phase_problem_sld} shows two hypothetical samples of the same total thickness,
consisting of one and two layers, deposited on a substrate with
$\rho_s = \scinot{2.07}{-6} \angstrom^{-2}$. Even though the SLD depth profiles are distinctly
different, the resultant reflectivity curves are virtually identical,
with only a slight discrepancy at the first minimum. This ambiguity,
known as \textit{the phase problem}, is
an example of the information lost in diffraction experiments due
to the inability of measuring the phase of the scattered wave
--Notice that equation \ref{equ:kine} involves a modulus-squared.

\begin{figure}
    \includegraphics[width=0.49\textwidth]{phase_problem_SLD.pdf} 
    \caption{Two different samples $\rho_{1,2}(z)$ deposited on the same substrate $\rho_s$ (top panel)
    are indistinguishable from one another by looking only at their reflectivity curve $R(q)$
    (bottom panel).}
    \label{fig:phase_problem_sld}
\end{figure}

An heuristic explanation for the phase problem
can be built in terms of the auto-correlation function of $\rmd\rho/\rmd z$, which gives a
real-space representation of the information contained in the reflectivity pattern:

\begin{equation}
    \begin{split}
        \acf_{\rho'}(z) & = \int_{-\infty}^{\infty} \rho'(z)^* \, \rho'(z + \xi) \, \rmd \xi \\
                        & = \int_{-\infty}^{\infty} \left| \widehat{\rho'}(q)  \right|^{2} \, \rm{e}^{i\,q\,z} \, \rmd q,
    \end{split}
    \label{equ:acf}
\end{equation}

\noindent where the hat $\widehat{[]}$ indicates a Fourier transform, the star $[]^*$ a complex conjugate
and the prime $[]'$ a derivative.

\begin{figure}
    \includegraphics[width=0.49\textwidth]{phase_problem_ACF.pdf}
    \caption{Top panel: the derivatives of the two hypothetical SLD profiles discussed, $\rho_{1,2}'(z)$, are
    scaled $\delta$ functions peaking at the interfaces between SLD layers. Bottom panel: In contrast,
    the autocorrelation function, $\acf_{\rho'}(z)$ is almost identical for both samples,
    peaking at exactly the same locations.}
    \label{fig:phase_problem_acf}
\end{figure}

As $\rho'(z)$ is purely real and in both examples is only composed of 
$\delta$ functions, the ACF can be obtained by direct inspection
(See Figure \ref{fig:phase_problem_acf} and Equation \ref{equ:acf}):

$\bullet$ At zero lag, there is always full correlation, thus the ACF will have 
a large positive peak at the origin for both samples.

$\bullet$ For both samples, the interference between the front and back interfaces, at $z = 0$ and
$z = 200 \angstrom$ respectively, translates into a pair of $\delta$ peaks
with opposite signs in $\rho'$. Thus, at a 
$\xi = \pm 200 \angstrom$ lag, the ACF must peak with negative amplitude.

$\bullet$ For the two-layer sample, an extra pair of peaks in the ACF could be expected
at lags of $\xi = \pm 100 \angstrom$; however, since there is a complete cancellation between
the positive and the negative contributions of the $\delta$ functions at $z = 0$ and $z = 200$,
the expected extra pair of peaks is absent.

\noindent As the sld profiles of both samples translate into the same ACF,
it is now not surprising the similarity between their reflectivity curves.

A deeper discussion on the phase problem and some techniques
that have been developed to tackle it are described in 
the work by \cite[Majkrzak et al.]{majkrzak_phase_2003}, who give an overview of several
phase-sensitive experimental methods and claim that any ambiguity in a SLD profile obtained
by these is a consequence only of the limited $q$ range over which reflectivities can be measured.
In a similar fashion, \cite{koutsioubas_model_2019} proposes a model-independent method that,
when applied to multiple solvent contrast data, should lead to reliable
reconstructions of the interfacial structure without the need for any a priori assumptions.

However, even when the phase problem can be theoretically alleviated, reflectivity curves
from different SLD profiles may still be close enough to one another to be indistinguishable
by finite-resolution experiments, effectively keeping the analysis of reflectometry profiles
in the realm of ill-posed non-invertible problems. As such, reflectometry data analysis demands
for advanced trial-and-error fitting techniques, the daily bread of experimenters to
which the following section is devoted.

%%%%
